% =============================================================
%  AI-Statement
% =============================================================

\section{Statement on the use of AI}\label{sec:ai}
%
AI tools were used extensively throughout the development of this project, primarily \textsc{ChatGPT} and \textsc{Claude}, 
with Claude playing a particularly significant role in several key aspects of the work.

\textbf{Code (notebooks and source files).} AI assistance was used for writing, editing, testing, and debugging the code. 
Some portions of the code were implemented from scratch by the AI based on my specifications, while other parts consisted 
of code I had already written that was subsequently refined, optimized, or corrected with AI assistance. All code was fully 
reviewed by me to ensure correctness and that I understand its functioning.

\textbf{Experiments and plots.} AI assistance was used in designing and setting up the experiments presented in this report, 
as well as in writing the code used to run them and to generate the plots. No figure included in this report was AI-generated 
as an image: every plot was produced by code (written with AI assistance) that I executed myself, and all figures reflect the 
actual results of the experiments I ran.

\textbf{Literature understanding and report consistency.} AI tools, especially Claude, were instrumental in achieving a 
deep understanding of the reference papers underlying this work, including clarifying derivations and concepts that were 
not initially clear. AI assistance was also used to identify ambiguities in both the report and the code, and to help 
maintain consistent notation throughout the report.

\textbf{Interpretation of results.} AI also contributed to the interpretation of the experimental results and to the 
discussion of the conclusions, in addition to helping express these observations clearly in the text.

\textbf{Report structure and writing.} The overall structure of the report was initially proposed by the AI; however, 
I judged it unsatisfactory and, through iterative prompt engineering, refined and reorganized it myself to emphasize 
the most important parts and to follow a coherent logical flow. Regarding the text, AI was used both to write paragraphs 
starting from my notes and drafts, and to edit and improve text I had already written, depending on the section. 
The entire report underwent a complete review by me.

\textbf{Overall.} The use of AI was substantial and constituted an important part of the overall workflow, particularly 
for understanding the theoretical foundations of the project and for designing and interpreting the experiments. 
Nonetheless, all AI-assisted material -- code, text, experiments, and explanations -- was reviewed, verified, and 
is fully understood by me, and I take full responsibility for the content of this report.
