% =============================================================
%  APPENDIX C — REWEIGHTING PIPELINE
% =============================================================

\section{Reweighting Pipeline}

% ------------------------------------------------------------------
\subsection{ResiDual: Implementation Details}
\label{app:residual_impl}
% ------------------------------------------------------------------

This section describes the full implementation of ResiDual (RD$^*$\xspace),
a learnable spectral adaptation integrated into a frozen CLAP model.
The pipeline consists of two sequential phases: a \emph{fitting pass}
that estimates a fixed PCA basis for each attention head
(\S\ref{app:fitting}), followed by an \emph{optimisation phase} in which
the per-head weight vectors are refined as free parameters via gradient
descent to maximise zero-shot classification accuracy (\S\ref{app:optim}).
The forward-pass reweighting applied at inference time is described in
\S\ref{app:forward_rw}.

Throughout this appendix we assume batch size $B = 1$ for notational
clarity; all operations extend straightforwardly to $B > 1$ by treating
the batch axis independently. We use the index conventions of
Appendix~\ref{app:architecture}: $\ell \in \{0,1,2,3\}$ is the stage
index, $b \in \{1,\ldots,B_\ell\}$ is the block index within stage
$\ell$, and $h \in \{1,\ldots,H_\ell\}$ is the head index. The
per-head feature dimension is $d_h = 24$ (constant across all stages),
$N_w^\ell = S_\ell^2 / M$ is the number of attention windows at stage
$\ell$, and $M = 64$ is the number of tokens per window. All symbols are
collected in Table~\ref{tab:notation}.

\paragraph{Spectral reweighting layers.}
For each target stage $\ell \in \mathcal{L}$, block $b$, and head $h$,
RD$^*$\xspace instantiates a \texttt{SpectralReweightingLayer} — a lightweight
\texttt{nn.Module} that carries two non-trainable buffers,
$\boldsymbol{\Phi}_{\ell,b,h} \in \mathbb{R}^{d_h \times d_h}$ and
$\boldsymbol{\mu}_{\ell,b,h} \in \mathbb{R}^{d_h}$, and a single
learnable parameter vector
$\boldsymbol{\lambda}_{\ell,b,h} \in \mathbb{R}^{d_h}$.
All such modules are collected in a nested \texttt{nn.ModuleDict}
indexed by $(\ell, b, h)$, constructed once during model initialisation
in \texttt{\_build\_spectral\_layers} from the stage dimensions and head
counts of the underlying HTS-AT configuration; they persist in the
model's \texttt{state\_dict} across saves and loads.
Each module exposes two methods: \texttt{fit\_pca}, which populates
$\boldsymbol{\Phi}_{\ell,b,h}$ and $\boldsymbol{\mu}_{\ell,b,h}$ from
data and initialises $\boldsymbol{\lambda}_{\ell,b,h}$
(§\ref{app:fitting}); and \texttt{forward}, which applies the
RD$^*$\xspace transform (§\ref{app:forward_rw}).

\paragraph{Module injection via forward hooks.}
RD$^*$\xspace injects its reweighting through PyTorch forward hooks.
For each target block $(\ell, b)$ with $\ell \in \mathcal{L}$, a
dedicated \texttt{AttentionHook} object is registered on the
corresponding \texttt{WindowAttention} module immediately before the
forward pass begins and removed as soon as that forward pass completes.
The hook is stateless: it holds only a reference to the corresponding
slice \texttt{spectral\_layers}$[\ell][b]$ of the module dict, and
operates in one of two modes controlled by a \texttt{collect\_for\_fitting}
flag. In \emph{collection mode}, it accumulates the raw per-head tensors
$\mathbf{H}_{\ell,b,h}$ without modifying the forward-pass output,
feeding the fitting pass of §\ref{app:fitting}. In \emph{reweighting
mode}, it calls \texttt{forward} on each \texttt{SpectralReweightingLayer}
and replaces $\mathbf{H}_{\ell,b,h}$ with the reweighted output
$\mathbf{H}^*_{\ell,b,h}$ before concatenation and output projection.
No structural modification of the original model is performed: all
pretrained parameters — the fused QKV projection $W^{QKV}_{\ell,b}$,
the output projection $W^O_{\ell,b}$ with bias $\mathbf{b}^O_{\ell,b}$,
and the relative position bias table $\mathbf{B}_{\ell,b,h}$ — remain
exactly as loaded from the checkpoint.

% ------------------------------------------------------------------
\subsubsection{Fitting Pass: PCA Basis Estimation}
\label{app:fitting}
% ------------------------------------------------------------------

\paragraph{Data collection.}
The fitting pass is run with gradients disabled and the model in
evaluation mode. For each audio clip (up to $N_{\max}$ clips), a
forward pass is executed with data collection enabled. The hook
registered at block $(\ell, b)$ intercepts, for every head $h$, the
tensor $\mathbf{H}_{\ell,b,h} \in \mathbb{R}^{N_w^\ell \times M \times d_h}$
and flattens it along the spatial axes to obtain a matrix of shape
$\mathbb{R}^{(N_w^\ell \cdot M) \times d_h}$, which is appended to an
accumulation buffer for the triplet $(\ell, b, h)$. After all clips
have been processed, the per-clip contributions are stacked to form the
\emph{fitting matrix}
\begin{equation}
    \widetilde{\mathbf{R}}_{\ell,b,h}
    \;\in\; \mathbb{R}^{N_{\mathrm{fit}} \times d_h},
    \qquad
    N_{\mathrm{fit}}
    \;=\; N_{\mathrm{clips}} \cdot N_w^\ell \cdot M,
    \label{eq:fitting_matrix}
\end{equation}
where $N_{\mathrm{clips}} \le N_{\max}$ is the number of processed clips.

\paragraph{PCA decomposition.}
Standard PCA is applied to $\widetilde{\mathbf{R}}_{\ell,b,h}$.
The empirical mean
\begin{equation}
    \boldsymbol{\mu}_{\ell,b,h}
    \;=\; \frac{1}{N_{\mathrm{fit}}}
          \sum_{i=1}^{N_{\mathrm{fit}}}
          \widetilde{\mathbf{R}}_{\ell,b,h}[i,:]
    \;\in\; \mathbb{R}^{d_h}
\end{equation}
is computed and subtracted, and the centred empirical covariance is
decomposed as
\begin{align}
    \mathbf{C}_{\ell,b,h}
    &= \frac{1}{N_{\mathrm{fit}} - 1}
       \Bigl(\widetilde{\mathbf{R}}_{\ell,b,h}
             - \mathbf{1}\boldsymbol{\mu}_{\ell,b,h}^\top\Bigr)^\top
       \Bigl(\widetilde{\mathbf{R}}_{\ell,b,h}
             - \mathbf{1}\boldsymbol{\mu}_{\ell,b,h}^\top\Bigr) \notag \\
    &= \boldsymbol{\Phi}_{\ell,b,h}\,
       \operatorname{diag}(\lambda^{\mathrm{ev}}_1, \ldots,
                            \lambda^{\mathrm{ev}}_{d_h})\,
       \boldsymbol{\Phi}_{\ell,b,h}^\top,
    \quad
    \lambda^{\mathrm{ev}}_1 \ge \cdots \ge
    \lambda^{\mathrm{ev}}_{d_h} \ge 0,
    \label{eq:pca_decomp}
\end{align}
where $\boldsymbol{\Phi}_{\ell,b,h} \in \mathbb{R}^{d_h \times d_h}$ is
the orthogonal matrix of eigenvectors (columns ordered by decreasing
eigenvalue) and $\lambda^{\mathrm{ev}}_j$ denotes the $j$-th eigenvalue
of $\mathbf{C}_{\ell,b,h}$. We use the superscript ``ev'' throughout to
distinguish these fixed quantities from the trainable weight vector
$\boldsymbol{\lambda}_{\ell,b,h}$.
Both $\boldsymbol{\Phi}_{\ell,b,h}$ and $\boldsymbol{\mu}_{\ell,b,h}$
are stored as non-trainable buffers; the accumulation buffers
$\widetilde{\mathbf{R}}_{\ell,b,h}$ are then cleared to free memory.

\paragraph{Task-relevant subspace and initialisation of
$\boldsymbol{\lambda}_{\ell,b,h}$.}
For each head $(\ell,b,h)$, the index $k_{\ell,b,h}$ defined in
Eq.~\eqref{eq:k_threshold} identifies the smallest number of principal
components that together explain at least a fraction $\rho_{\mathrm{var}}$
of the total variance, providing a per-head estimate of the effective
dimensionality of the task-relevant subspace.
The learnable weight vector
$\boldsymbol{\lambda}_{\ell,b,h} \in \mathbb{R}^{d_h}$
is initialised component-wise as
\begin{equation}
    \lambda_j^{(0)}
    \;=\; \begin{cases}
        1 & j \le k_{\ell,b,h}, \\
        0 & j > k_{\ell,b,h},
    \end{cases}
    \label{eq:lambda_init}
\end{equation}
corresponding to the identity map restricted to the task-relevant
subspace and complete suppression of the remaining noise components.
At this initialisation point, RD$^*$\xspace therefore acts as a
projection-and-reconstruction operator that retains the top-$k_{\ell,b,h}$
PCA directions unchanged and zeros out the remainder.

% ------------------------------------------------------------------
\subsubsection{Optimisation of $\boldsymbol{\lambda}$ via Zero-Shot Accuracy}
\label{app:optim}
% ------------------------------------------------------------------

After the fitting pass has fixed the PCA bases
$\{\boldsymbol{\Phi}_{\ell,b,h}\}$ and initialised
$\{\boldsymbol{\lambda}_{\ell,b,h}\}$ via Eq.~\eqref{eq:lambda_init},
the second phase optimises the weight vectors to maximise zero-shot
classification accuracy on a held-out validation set.

\paragraph{Parameter isolation.}
All model parameters except the weight vectors
$\{\boldsymbol{\lambda}_{\ell,b,h}\}_{\ell \in \mathcal{L},\,b,\,h}$
are frozen by setting \texttt{requires\_grad = False}. The only degrees
of freedom during optimisation are therefore the
$\sum_{\ell \in \mathcal{L}} B_\ell \cdot H_\ell$ weight vectors, each
of dimension $d_h = 24$, that control the per-component gain of each
attention head in the target stages $\mathcal{L}$. The CLAP audio
projection head $P : \mathbb{R}^{D_3} \to \mathbb{R}^{d_{\mathrm{proj}}}$
is kept frozen but is included in the computational graph so that
gradients flow from the similarity space back to
$\boldsymbol{\lambda}_{\ell,b,h}$.

\paragraph{Objective.}
Given a set of audio clips with ground-truth class labels
$y \in \{1,\ldots,C\}$ and pre-computed $\ell_2$-normalised text
embeddings $\mathbf{t}_c \in \mathbb{R}^{d_{\mathrm{proj}}}$ for each
class $c$, the optimisation objective is the cross-entropy loss over
zero-shot logits:
\begin{equation}
    \mathcal{J}(\boldsymbol{\lambda})
    \;=\;
    -\,\mathbb{E}_{(x,\,y)}\!\left[
        \log \frac{
            \exp\!\bigl(\hat{\mathbf{a}}(x)^\top \mathbf{t}_y\bigr)
        }{
            \displaystyle\sum_{c=1}^{C}
            \exp\!\bigl(\hat{\mathbf{a}}(x)^\top \mathbf{t}_c\bigr)
        }
    \right],
    \label{eq:optim_obj}
\end{equation}
where $\hat{\mathbf{a}}(x) \in \mathbb{R}^{d_{\mathrm{proj}}}$ is the
$\ell_2$-normalised audio embedding produced by $P$ applied to the
spatially averaged Stage-3 output of HTS-AT after the RD$^*$\xspace
transforms have been applied (cf.\ Table~\ref{tab:notation}).
Minimising $\mathcal{J}$ is equivalent to maximising zero-shot
classification accuracy in expectation.

\paragraph{Optimiser and schedule.}
The weight vectors $\{\boldsymbol{\lambda}_{\ell,b,h}\}$ are updated
with Schedule-Free Adam~\cite{defazio2024schedulefree} at learning rate
$\eta$. This optimiser eliminates the need for a separate
learning-rate schedule while matching or exceeding the performance of
Adam with cosine decay on this class of fine-tuning tasks.
When the \texttt{schedulefree} package is unavailable, standard Adam
is used as a fallback.

\paragraph{Early stopping and best-$\boldsymbol{\lambda}$ restoration.}
At the end of each epoch, zero-shot accuracy on the validation set is
evaluated. A snapshot of the current
$\{\boldsymbol{\lambda}_{\ell,b,h}\}$ is saved whenever a new accuracy
maximum is achieved. Training halts early if no improvement is observed
for $p$ consecutive epochs; at termination, the best snapshot is
restored, guaranteeing that the final weight vectors correspond to the
configuration achieving the highest observed validation accuracy.

\paragraph{Hyperparameter summary.}
Table~\ref{tab:residual_hparams} collects all hyperparameters introduced
by RD$^*$\xspace together with the default values used in all experiments.

\begin{table}[h]
\centering
\caption{RD$^*$\xspace hyperparameters and default values.}
\label{tab:residual_hparams}
\begin{tabular}{lll}
\toprule
\textbf{Symbol} & \textbf{Description} & \textbf{Default} \\
\midrule
$\mathcal{L}$
    & Target stage indices
    & $\{1, 2, 3\}$ \\
$\rho_{\mathrm{var}}$
    & Variance threshold for $k_{\ell,b,h}$
      (Eq.~\eqref{eq:k_threshold})
    & $0.95$ \\
$\eta$
    & Learning rate (Schedule-Free Adam)
    & $10^{-2}$ \\
$p$
    & Early-stopping patience (epochs)
    & $5$ \\
$T_{\max}$
    & Maximum optimisation epochs
    & $30$ \\
$N_{\max}$
    & Maximum fitting clips
    & $10\,000$ \\
\bottomrule
\end{tabular}
\end{table}

% ------------------------------------------------------------------
\paragraph{Remark on hook-based extraction.}
% ------------------------------------------------------------------

The \texttt{AttentionHook} mechanism described in \S\ref{sec:aggregation}
for the dimensionality analysis is the same mechanism through which
RD$^*$\xspace applies reweighting at inference time. The hook operates in
two distinct modes controlled by an internal \texttt{collect\_for\_fitting}
flag: in \emph{collection mode} it accumulates raw head outputs
$\mathbf{H}_{\ell,b,h}$ for PCA estimation without modifying the
forward-pass output; in \emph{reweighting mode} it replaces
$\mathbf{H}_{\ell,b,h}$ with $\mathbf{H}^*_{\ell,b,h}$
(Eq.~\eqref{eq:reweighted_head}) before the concatenation and output
projection of Eq.~\eqref{eq:wmsa_out}. In both modes the hooks are
\emph{temporary}: they are installed at the start of each call to the
instrumented forward function and removed immediately upon its completion,
so the underlying \texttt{WindowAttention} modules are never
structurally modified.

% ------------------------------------------------------------------
\subsubsection{Forward Pass with Spectral Reweighting}
\label{app:forward_rw}
% ------------------------------------------------------------------

At inference time, the hook at block $(\ell, b)$ intercepts the raw head
outputs $\mathbf{H}_{\ell,b,h} \in \mathbb{R}^{N_w^\ell \times M \times d_h}$
(cf.\ Eq.~\eqref{eq:raw_head}) and applies the following sequence of
operations independently for each head $h$.

\textbf{Step 1 — Centering.}
\begin{equation}
    \widetilde{\mathbf{H}}_{\ell,b,h}
    \;=\; \mathbf{H}_{\ell,b,h} - \boldsymbol{\mu}_{\ell,b,h}
    \;\in\; \mathbb{R}^{N_w^\ell \times M \times d_h},
\end{equation}
where $\boldsymbol{\mu}_{\ell,b,h} \in \mathbb{R}^{d_h}$ is broadcast
over the first two axes.

\textbf{Step 2 — Projection onto the full PCA basis.}
\begin{equation}
    \mathbf{Z}_{\ell,b,h}
    \;=\; \widetilde{\mathbf{H}}_{\ell,b,h}\,\boldsymbol{\Phi}_{\ell,b,h}
    \;\in\; \mathbb{R}^{N_w^\ell \times M \times d_h}.
\end{equation}
Unlike fixed-weight variants that truncate to $k_{\ell,b,h} < d_h$
components, RD$^*$\xspace projects onto the \emph{full} basis
$\boldsymbol{\Phi}_{\ell,b,h} \in \mathbb{R}^{d_h \times d_h}$,
delegating the effective truncation to the learned weights.

\textbf{Step 3 — Per-component rescaling.}
\begin{equation}
    \widehat{\mathbf{Z}}_{\ell,b,h}
    \;=\; \mathbf{Z}_{\ell,b,h}
          \odot \boldsymbol{\lambda}_{\ell,b,h}^\top
    \;\in\; \mathbb{R}^{N_w^\ell \times M \times d_h},
    \label{eq:rd_reweight}
\end{equation}
where $\boldsymbol{\lambda}_{\ell,b,h}^\top \in \mathbb{R}^{d_h}$ is
broadcast over the first two axes. Because $\boldsymbol{\lambda}_{\ell,b,h}$
is a free real-valued parameter, each entry $\lambda_j$ can amplify
($|\lambda_j|>1$), attenuate ($|\lambda_j|\in(0,1)$), suppress
($\lambda_j=0$), or invert ($\lambda_j<0$) the corresponding principal
component.

\textbf{Step 4 — Reconstruction in the original space.}
\begin{equation}
    \mathbf{H}^*_{\ell,b,h}
    \;=\; \widehat{\mathbf{Z}}_{\ell,b,h}\,\boldsymbol{\Phi}_{\ell,b,h}^\top
    \;\in\; \mathbb{R}^{N_w^\ell \times M \times d_h}.
    \label{eq:reweighted_head}
\end{equation}
The empirical mean $\boldsymbol{\mu}_{\ell,b,h}$ is \emph{not} restored
after reconstruction. Since $\mathbf{H}^*_{\ell,b,h}$ is immediately
passed to the output projection $W^O_{\ell,b}$ (cf.\
Eq.~\eqref{eq:wmsa_out}), which carries its own learnable bias
$\mathbf{b}^O_{\ell,b} \in \mathbb{R}^{D_\ell}$, re-adding
$\boldsymbol{\mu}_{\ell,b,h}$ would introduce a systematic,
data-dependent shift in the residual stream that is already absorbed by
$\mathbf{b}^O_{\ell,b}$.

Combining Steps~1--4, the full RD$^*$\xspace transform reads:
\begin{equation}
    \mathrm{RD}^*\!\left(\mathbf{H}_{\ell,b,h},\,
                          \boldsymbol{\lambda}_{\ell,b,h}\right)
    \;=\;
    \left(\mathbf{H}_{\ell,b,h} - \boldsymbol{\mu}_{\ell,b,h}\right)
    \boldsymbol{\Phi}_{\ell,b,h}\,
    \operatorname{diag}(\boldsymbol{\lambda}_{\ell,b,h})\,
    \boldsymbol{\Phi}_{\ell,b,h}^\top
    \;\in\; \mathbb{R}^{N_w^\ell \times M \times d_h}.
    \label{eq:rd_star}
\end{equation}
The tensor $\mathbf{H}^*_{\ell,b,h}$ replaces $\mathbf{H}_{\ell,b,h}$
in Eq.~\eqref{eq:wmsa_out}, so the output projection $W^O_{\ell,b}$
and the residual-stream update of Eq.~\eqref{eq:residual_attn} proceed
unchanged. During optimisation, gradients of any downstream loss flow
through $\mathrm{RD}^*$ and update only
$\boldsymbol{\lambda}_{\ell,b,h}$; the basis
$\boldsymbol{\Phi}_{\ell,b,h}$, the mean $\boldsymbol{\mu}_{\ell,b,h}$,
and all original model parameters are held fixed.
